---
title: "Raptor Toxicant Preliminary Analysis"
author: "Pat Dumandan"
date: "12/10/2020"
output: html_document
---
```{r setup, include=FALSE, message=FALSE, echo=FALSE, warning=FALSE}
require(dplyr)
require(rstan)
require(rstanarm)

options(mc.cores = parallel::detectCores())

tox=read.csv("prax_sheet.csv")
str(tox)
```

### Data Cleaning
```{r, message=FALSE, echo=FALSE, warning=FALSE}
tox_conc=tox%>%
rename(proportion=OR.if.given..proportion.exposed,concentration=concentration..ND.for.non.detects. )%>%
filter(!is.na(concentration))%>%
arrange(ID)

unique(tox_conc$toxicant.group)
```
*Q for Vince: 1) what are PFAs, Os, unknowns?, 2) how is metal and heavy metal dfferent? *

### Analysis Plan  

#### Part 1: Toxicant-specific  
*assess how toxicant concentrations vary per country, species, feeding trait, etc.*  

Steps:  
  * identify the most common/ most studied specific toxicant per main toxicant group (also identify what sample type collected)   
  * standardize values across all studies
  * create forest plot of top toxicant specific concentrations  
  * create mixed-effects model with intercept allowed to vary per study, species, country   
  

#### Example: Heavy Metals

```{r, message=FALSE, echo=FALSE, warning=FALSE}
met=c("heavy metals", "metal")
metal=tox_conc%>%
  filter(toxicant.group %in% met)%>%
  group_by (toxicant.specific, sample.type)%>%
  summarise(count=n())%>%
  arrange(count)
```
**Pb (Lead) in liver is most well-studied**  

```{r, message=FALSE, echo=FALSE, warning=FALSE}

```